---
title: "Working with LAGOS"
author: "Joseph Stachelek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with LAGOS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The following document presents two analyses to demonstrate use of the `LAGOS` package.

To begin, we load LAGOS data with `lagos_load`:

```{r lagos_load}
library(LAGOS)

dt <- lagos_load("1.087.1")
```

## Boxplot of average chlorophyll by state

The first analysis involves contructing a boxplot of average lake chlorophyll by state. To begin, we join the nutrient water quality table (`epi.nutr`) to the location information for each lake in the `locus` table. Next, we join the output to the `state` table to obtain state abbreviations that correspond to each lake. Finally, we filter the dataset to exclude any lakes that do not fall within a state. 

```{r avg_chl_by_state, message=FALSE, results = 'hide', warning = FALSE}
library(dplyr)
library(ggplot2)

lg <- left_join(dt$epi.nutr, dt$locus)
lg <- left_join(lg, dt$state)
lg <- group_by(lg, state)
lg <- filter(lg, !is.na(state))
```

Now, the `levels` of the `state` factor column are in alphabetical order by default. We need to reorder the factor levels from East to West in preparation for plotting.

```{r }
levels(lg$state)
state_easting <- summarize(lg, 
                    mean_easting = mean(nhd_long, na.rm = TRUE))
state_easting <- arrange(state_easting, mean_easting)

lg$state <- factor(lg$state, levels = state_easting$state)

```

Finally, we contruct a boxplot with `ggplot`:

```{r warning=FALSE}
ggplot(lg) + 
  geom_boxplot(aes(x = state, y = chla), outlier.shape = NA) + 
  ylim(c(0, 80)) + ylab("Chlorophyll a (ug/L)")

```

## Map of average chlorophyll in Pennsylvania

The second analysis involves contructing a map of average lake chlorophyll concentration in Pennsylvania lakes. To begin, we use the `maps` and `sf` package to obtain a geographic outline of the state.  

```{r pa_chl_map, message=FALSE, results = 'hide'}
library(maps)
library(sf)

states <- st_as_sf(map("state", plot = FALSE, fill = TRUE))
pa <- filter(states, ID == "pennsylvania")
```

Next, we filter the output from the previous analysis to only include PA lakes and calculate mean chlorophyll for each lake: 

```{r }
pa_chl <- filter(lg, state == "PA")
pa_chl <- group_by(pa_chl, lagoslakeid)
pa_chl <- summarize(pa_chl, 
                      mean_chl  = mean(chla, na.rm = TRUE), 
                      mean_long = mean(nhd_long, na.rm = TRUE), 
                      mean_lat  = mean(nhd_lat, na.rm = TRUE))
pa_chl <- filter(pa_chl, !is.na(mean_chl))
```

Finally, we turn our `data.frame` into an sf geospatial object using `st_as_sf` before constructing a map with `ggplot` and `geom_sf`.

```{r }
pa_chl <- st_as_sf(pa_chl, coords = c("mean_long", "mean_lat"), crs = 4326)

ggplot() + 
  geom_sf(data = pa) +
  geom_sf(data = pa_chl, aes(color = log(mean_chl))) + 
  theme_bw() + scale_color_gradient(low = "white", high = "darkgreen") +
  labs(color = "log(Chlorophyll a (ug/L))")

```
