---
title: "Working with LAGOSNE"
author: "Joseph Stachelek"
date: "`r Sys.Date()`"
output: 
  html_document:
      fig_caption: yes
---

The following document demonstrates use of the `LAGOSNE` package with two analyses:

Before we can begin, we must load LAGOSNE data with `lagosne_load`:

```{r lagos_load, eval=FALSE}
library(LAGOSNE)
dt <- lagosne_load()
```

```{r lagos_load_cached, echo=FALSE}
library(LAGOSNE)
dt <- readRDS(system.file("lagos_test_subset.rds", package = "LAGOSNE"))
```

## Boxplot of average chlorophyll by state

The first analysis involves constructing a boxplot of average lake chlorophyll by state. To begin, we join the nutrient water quality table (`epi.nutr`) to the location information for each lake in the `locus` table. Next, we join the output to the `state` table to obtain state abbreviations that correspond to each lake. Finally, we filter the dataset to exclude any lakes that do not fall within a state. 

```{r avg_chl_by_state, message=FALSE, results = 'hide', warning = FALSE, eval=TRUE}
library(dplyr)
library(ggplot2)

lg <- left_join(dt$epi_nutr, dt$locus)
lg <- left_join(lg, dt$state)
lg <- group_by(lg, state)
lg <- filter(lg, !is.na(state))
```

Now, the `state` column is of `character` type so will be in alphabetical order by default. We need to cast this column as a factor and reorder its levels from East to West in preparation for plotting.

```{r eval=TRUE}
state_easting <- summarize(lg, 
                    mean_easting = mean(nhd_long, na.rm = TRUE))
state_easting <- arrange(state_easting, mean_easting)

lg$state <- factor(lg$state, levels = state_easting$state)
```

Finally, we contruct a boxplot with `ggplot`:

```{r warning=FALSE, eval=TRUE}
ggplot(lg) + 
  geom_boxplot(aes(x = state, y = chla), outlier.shape = NA) + 
  ylim(c(0, 80)) + ylab("Chlorophyll a (ug/L)")
```

## Map of average chlorophyll in Pennsylvania

The second analysis involves constructing a map of average lake chlorophyll concentration in Pennsylvania lakes. To begin, we use the `maps` and `sf` package to obtain a geographic outline of the state.  

```{r pa_chl_map, message=FALSE, results = 'hide', eval=FALSE}
library(maps)
library(sf)

states <- st_as_sf(map("state", plot = FALSE, fill = TRUE))
pa     <- filter(states, ID == "pennsylvania")
```

Next, we filter the output from the previous analysis to only include PA lakes and calculate mean chlorophyll for each lake: 

```{r eval=FALSE}
pa_chl <- filter(lg, state == "PA")
pa_chl <- group_by(pa_chl, lagoslakeid)
pa_chl <- summarize(pa_chl, 
                      mean_chl  = mean(chla, na.rm = TRUE), 
                      mean_long = mean(nhd_long, na.rm = TRUE), 
                      mean_lat  = mean(nhd_lat, na.rm = TRUE))
pa_chl <- filter(pa_chl, !is.na(mean_chl))
```

Finally, we turn our `data.frame` into an sf geospatial object using `st_as_sf` before constructing a map with `ggplot` and `geom_sf`.

```{r eval=FALSE}
pa_chl <- st_as_sf(pa_chl, coords = c("mean_long", "mean_lat"), crs = 4326)

ggplot() + 
  geom_sf(data = pa) +
  geom_sf(data = pa_chl, aes(color = log(mean_chl))) + 
  theme_bw() + scale_color_gradient(low = "white", high = "darkgreen") +
  labs(color = "log(Chlorophyll a (ug/L))")
```
